Parsing Postfix log files

### 2007-06-11 doc/logparser.tex
logparser.tex: * Documented SQL schema. * The TOC is now linked for easy navigation, and there are bookmarks in Acrobat too. * Misc cleanups. Makefile: produce a PDF automatically.

% Better typesetting of URLs.
\usepackage{url}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
\usepackage[section,numbib,nottoc]{tocbibind}

@@

@@

@@

    \item Checking Sender Policy Framework
        records~\cite{openspf},~\cite{wikipedia-spf}.  SPF records specify
        which mail servers are allowed to send mail claiming to be from a
        particular domain.  The intention is to reduce spam from faked
        sender addresses, backscatter~\cite{postfix-backscatter} and
        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of SMTP\@.

@@

@@

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a ``heads up'' for potential trouble spots.\/}
        \url{http://jimsun.linxnet.com/postfix_contrib.html}

@@

\subsection{Ratio of file size to execution time and number of lines to
execution time}
\label{execution time vs file size vs number lines factor}
\includegraphics{plot-normal-filesize-numlines-factor.ps}

@@

\subsection{Percentage increase of shuffled over normal and reversed over
normal}
\label{normal vs shuffled vs reversed ordering factor}
\includegraphics{plot-normal-shuffle-reverse-factor.ps}

>>>>

% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib,nottoc]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.
\usepackage{hyperref}
\hypersetup{%
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final,
    pdfborder   = {0 0 0},
    breaklinks,
}

@@

XXX LOTS MORE REFERENCES

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and cause the correct action to be executed; they also have several
fields which aid the user in understanding what each rule parses.  The
rules are described in detail in section~\ref{rules} and the fields in
particular in section~\ref{rule attributes}, and that discussion won't be
repeated here.  The data saved from parsing the logs is divided into two
tables: connections and results.

\subsubsection{Connections table}

Every connection and/or mail will have a single entry in the connections
table containing the following fields:

    \item [id] This field uniquely identifies the row.

    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
        local server when receiving mail, the remote server when sending
        mail.

    \item [server\_hostname] The hostname of the server, it will be
        \textit{unknown} if the IP address could not be resolved to a
        hostname via DNS.

    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
        server when receiving mail, the local server when sending mail.

    \item [client\_hostname] The hostname of the client, it will be
        \textit{unknown} if the IP address could not be resolved to a
        hostname via DNS.

    \item [helo] The hostname used in the HELO command.  The HELO
        occasionally changes during a connection, presumably because spam
        or virus senders think it's a good idea.  By default Postfix only
        logs this when it rejects an SMTP command, but it is trivial to log
        this for every connection:

        \texttt{smtpd\_helo\_restrictions = \newline
        \hspace*{2em} check\_helo\_access regexp:/etc/postfix/log\_helo.regex}

        Where \texttt{/etc/postfix/log\_helo.regex} contains:

        \texttt{/./ WARN Logging HELO for diagnostic purposes}

    \item [queueid] The queueid of the mail, or \textit{NOQUEUE} if a mail
        wasn't accepted.

    \item [start] The timestamp of the first log line, in seconds since the
        epoch.

    \item [end] The timestamp of the last log line, in seconds since the
        epoch.

\subsubsection{Results table}

Every log line corresponding to Postfix performing an action other than
INFO will have an entry in the results table, e.g. rejecting an SMTP
command, delivering a mail, or bouncing a mail.  Any log line where the
Postfix action is INFO is not interesting in and of itself, but provides
additional information which will be saved with other results.

    \item [connection\_id] A reference to the entry in the connections
        table this results entry belongs to.

    \item [rule\_id] A reference to the entry in the rules table which
        matched the log line and created this result.

    \item [warning] Postfix can log a warning instead of rejecting an SMTP
        command - a facility that is quite useful for testing new
        restrictions.  This field will be 1 if the log line parsed was a
        warning rather than a real rejection, or 0 for a real rejection or
        any other result.

    \item [smtp\_code] The SMTP code extracted from the log line.  In
        general an SMTP code is only present for a rejection or final
        delivery; results missing an SMTP code will duplicate the SMTP code
        of other results.  Some final delivery log lines don't contain an
        SMTP code: in those cases the code is faked based on the success or
        failure represented by the log line.

    \item [sender] The sender's email address.  This can change during one
        single connection, when the connection is reused to send multiple
        mails.

    \item [recipient] The recipient address.

    \item [message\_id] The message-id of the accepted mail, or NULL if no
        mail was accepted.

    \item [data] A field available for anything not covered by other
        fields, e.g. the rejection message from an RBL.

    \item [timestamp] The time the line was logged, in seconds since the
        epoch.

@@

    \item Checking Sender Policy Framework records~\cite{openspf,
        wikipedia-spf}.  SPF records specify which mail servers are allowed
        to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of SMTP\@.

@@

\label{rule attributes}

@@

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a ``heads up'' for potential trouble spots.\/}
        \newline \url{http://jimsun.linxnet.com/postfix_contrib.html}

@@

\subsection{Ratio of file size and number of lines to execution time}
\label{execution time vs file size vs number lines factor}
\includegraphics{plot-normal-filesize-numlines-factor.ps}

@@

\subsection{Percentage increase of shuffled and reversed over normal}
\label{normal vs shuffled vs reversed ordering factor}
\includegraphics{plot-normal-shuffle-reverse-factor.ps}
